Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-08-16 prop_letter.tex
From: Hanson, Andrew J. Sent: Tuesday, August 15, 2017 18:48 To: Tai, Yu-Tsung <yuttai@indiana.edu> Cc: Ortiz, Gerardo <ortizg@indiana.edu>; Sabry, Amr A. <sabry@indiana.edu> Subject: My version of section I
Hi all -
    I am afraid of messing things up, so I just took out a copy of the file, and then edited in a separate subfile -- here is what I came up with for a rephrasing of Section I, which is all I have time for now.  I tried to maintain as much as possible in unaltered lines to make  'xxdiff'  more relevant.
   Are we meeting again on Thursday, or when?

   -Andy

There are long-standing debates in the foundations of quantum
mechanics regarding the tension between finite precision measurement
and contextuality~\cite{BarrettKent2004,Appleby_2005}, which refers
to the fact that the results of measuring compatible sets of
observables depend on whether they are measured together or
separately.  For example, Meyer~\cite{PhysRevLett.83.3751} and
Mermin~\cite{Mermin1999} disagree on whether finite-precision
measurements invalidate the spirit of the Kochen-Specker
theorem~\cite{kochenspecker1967,peres1995quantum}, which is a
mathematical statement about contextuality.  The Kochen-Specker
theorem asserts that, in a Hilbert space of dimension $d \ge 3$, it is
impossible to associate definite measured values, $\mu(P_i)= 0$ or
$1$, with every projection operator~$P_i$, in such a way that, if a
set of commuting $P_i$ satisfies $\sum_i P_i = 1$, then
$\sum_i \mu(P_i) = 1$. In contrast, Meyer showed that one {\it can\/}
assign definite values $0$ or $1$ to the measurement outcome
$\mu(P_i)$ when the Hilbert space is defined over the field of
rational numbers, therefore nullifying the
theorem~\citep{PhysRevLett.83.3751}.  However, Mermin then argued
that measurement outcomes must depend smoothly on slight changes in
the experimental configuration. Mermin therefore claimed that the
impact of Meyer's negative result is
``unsupportable''~\citep{Mermin1999}.  Discussion has continued on
whether contextuality is consistent with experimental
outcomes~\citep{Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.

A fundamental question is how to address this controversy by
developing mathematical theories of quantum mechanics that are
intrinsically, rather than only implicitly, consistent with the
resources available for the realistic accuracy of an actual
measurement. In this paper, we extend our previous work on the
foundations of computability in quantum physics by exploring the
application of interval-valued probability measures (IVPMs) to
achieving a coherent formulation of finite-resource quantum mechanics.
To this end, we develop a mathematical framework that includes the
uncertainties of finite precision and imperfections in the quantum
measurement process. In particular, we quantify the concept of
imperfect quantum measurement and its impact on the interpretation of
the Kochen-Specker and Gleason theorems and their implications for the
foundations of quantum mechanics.  Our first observation is that
essential objectives of our program to achieve a computable theory of
quantum mechanics and finite-resource measurement, attempted
previously with computable number
systems~\cite{usat,geometry2013apsrev4,DQT2014}, may be achievable by
extending classical IVPMs~\cite{JamisonLodwick2004} to the quantum
domain.  We thus begin by axiomatizing a quantum interval-valued
probability measure (QIVPM) framework.

The effect of finite precision and imperfections on contextuality and
the Kochen-Specker theorem is shown to have a resolution in terms of
our QIVPM framework.  To proceed, we use a formulation of the
``Mermin-Peres square''~\cite{peres1995quantum,Mermin1990Simple}
that uses the expectation value of a physical observable with respect
to a QIVPM, where the interval-valued expectation value is consistent
with the expectation value of the same observable in conventional
quantum mechanics with infinite resources. We show that, as long as
errors and uncertainties remain smaller than a particular value, the
Kochen-Specker theorem remains valid. In other words, we established a
rigorous bound for its validity and, therefore, the limits for
physical reality to remain contextual.

A key aspect of a quantum measurement is its statistical nature, as
reflected by application of the Born rule to determine the probability
of a given measurement
outcome~\citep{Born1983,peres1995quantum,544199}. The Gleason theorem
establishes that there is no alternative to the Born rule by
demonstrating, using reasonable continuity arguments in Hilbert spaces
of dimension $d\ge3$, the existence of a unique state $\rho$
consistent with the statistical predictions computed from that
rule~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.  It is
then legitimate to question what happens when infinite precision
cannot be achieved. We will show that, if the effect of finite
precision is incorporated, while a QIVPM might not be consistent with
a unique state~$\rho$ on all projectors defined on a Hilbert space
$\Hilb$ of dimension $d\ge3$, there is a mathematically precise sense
in which one recovers the original Gleason theorem asymptotically.
Specifically, if for bounded resources the QIVPM is induced by some
states~$\rho'$, those states will converge to a unique state $\rho$
when the size of the intervals characterizing the uncertainties tends
to zero.

The paper is organized as follows.  In Sec.\ \ref{sec:fuzzy} we begin
by introducing the foundations of quantum probability space and fuzzy
measurement.  We then move on to introduce the concept of quantum
interval valued probability in Sec.\ \ref{sec:Interval-Uncertainty}.
This QIVPM will provide the mathematical framework to quantify the
impact of finite precision measurements on quantum mechanics.  For
instance, in Sec.\ \ref{sec:Kochen-Specker}, we quantify the domain of
validity of a contextual measurement, thus addressing the conditions
under which the Kochen-Specker theorem applies.  This provides a way
of resolving the Meyer-Mermin debate, which is the main result of our
Thm~\ref{cor:Kochen-Specker-IVPM}.  In Sec.\ \ref{sec:Gleason}, we
study Gleason's theorem in the context of imprecise measurements,
concluding that QIVPM provides a framework with quantitative bounds in
which Gleason's theorem, while formally invalid in a universe with
bounded resources, holds asymptotically in a mathematically precise
way.  Finally, we present our conclusions in
Sec.~\ref{sec:Conclusion}.

>>>>

There are long-standing debates in the foundations of quantum
mechanics regarding the tension between finite precision measurement
and contextuality~\cite{BarrettKent2004,Appleby_2005}. The outcome
of a theory that is contextual depends upon
whether  compatible sets of observables are measured together or
separately; a non-contextual theory gives the same results in either
case.  A classic example of the varying opinions on the impact of
imprecision is the disagreement between
Meyer~\cite{PhysRevLett.83.3751} and 
Mermin~\cite{Mermin1999}  on whether finite-precision
measurements invalidate the spirit of the Kochen-Specker
theorem~\cite{kochenspecker1967,peres1995quantum}.  The Kochen-Specker  
theorem is in essence a
mathematical statement about contextuality, asserting that
in a Hilbert space of dimension $d \ge 3$, it is
impossible to associate definite measured values, $\mu(P_i)= 0$ or
$1$, with every projection operator~$P_i$, in such a way that, if a
set of commuting $P_i$ satisfies $\sum_i P_i = 1$, then
$\sum_i \mu(P_i) = 1$.   Meyer showed that one {\it can\/}
assign definite values $0$ or $1$ to the measurement outcome
$\mu(P_i)$ when the Hilbert space is defined over the field of
rational numbers, therefore nullifying the
theorem~\citep{PhysRevLett.83.3751}.  However, Mermin then argued
that measurement outcomes must depend smoothly on slight changes in
the experimental configuration, leading him to assert that the
impact of Meyer's negative result is
``unsupportable''~\citep{Mermin1999}.  The question of
whether contextuality is consistent with experimental
outcomes continues to be an active topic of discussion~\citep{Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.

The question of how to address this controversy is therefore a
fundamental problem.  How does one
develop mathematical theories of quantum mechanics that are
intrinsically, rather than only implicitly, consistent with the
resources available for the realistic accuracy of an actual
measurement?   In this paper, we extend our previous work on the
foundations of computability in quantum physics by exploring the
application of interval-valued probability measures (IVPMs) to
achieving a coherent formulation of finite-resource quantum mechanics.
From this starting point, we develop a mathematical framework that includes the
uncertainties of finite precision and imperfections in the quantum
measurement process.  In particular, we are able to quantify the concept of
imperfect quantum measurement and its impact on the interpretation of
the Kochen-Specker and Gleason theorems and their implications for the
foundations of quantum mechanics.  We have reason to believe that
essential objectives of our program to achieve a computable theory of
quantum mechanics and finite-resource measurement, attempted
previously with computable number
systems~\cite{usat,geometry2013apsrev4,DQT2014}, may be achievable by
extending classical IVPMs~\cite{JamisonLodwick2004} to the quantum
domain.  We thus begin by axiomatizing a quantum interval-valued
probability measure (QIVPM) framework.

A principal result is that recasting the Kochen-Specker theorem in
terms of ``$\delta$-determinism'', with $\delta$ denoting a
\emph{range} of outcomes, reducing to the purely deterministic
Kochen-Specker case for $\delta=0$, can resolve the question of effect
of finite precision and imperfections on contextuality.  A remarkable
consequence of our analysis is that not only does the
quantum-mechanical requirement for contextuality continue to hold,
thus maintaining the Kochen-Specker result for a \emph{non-vanishing}
range of $\delta$, but that at a \emph{certain fixed value} of
$\delta$, there is a sharp transition, parallel in its consequences
to, say, a phase transition, to \emph{non-contextuality}; that is,
quantum mechanics suddenly fails to hold, and all variables can be
specified simultaneously, turning the theory effectively into a
classical theory that can be specified consistently by hidden
variables.  These results provide a new insight into the
Meyer-Mermin debate by presenting a theory in which there is a
\emph{parameter} interpolating between what appeared previously to be
irreconcilable interpretations of contextuality.

%   To proceed, we use a formulation of the
% ``Mermin-Peres square''~\cite{peres1995quantum,Mermin1990Simple}
% that uses the expectation value of a physical observable with respect
% to a QIVPM, where the interval-valued expectation value is consistent
% with the expectation value of the same observable in conventional
% quantum mechanics with infinite resources. We show that, as long as
% errors and uncertainties remain smaller than a particular value, the
% Kochen-Specker theorem remains valid. In other words, we established a
% rigorous bound for its validity and, therefore, the limits for
% physical reality to remain contextual.

We also investigate, though with less insight-generating results, the
second key aspect of imprecise quantum measurement, which is its
statistical nature as reflected in the application of the Born rule
determining the probability of a given measurement
outcome~\citep{Born1983,peres1995quantum,544199}.  Here, the Gleason
theorem establishes that there is no alternative to the Born rule by
demonstrating, using reasonable continuity arguments in Hilbert spaces
of dimension $d\ge3$, the existence of a unique state $\rho$
consistent with the statistical predictions computed from that
rule~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.  Our
question then concerns what happens when infinite precision cannot be
achieved.  We are able to show that, while a QIVPM incorporating the
effects of finite precision might not be consistent with Gleason's
unique state~$\rho$ on all projectors defined on a Hilbert space
$\Hilb$ of dimension $d\ge3$, there is a mathematically precise sense
in which one recovers the original Gleason theorem asymptotically.
Specifically, if a set of QIVPMs representing bounded resources
correspond individually to at least one state's~$\rho'$, those states
will converge to a unique state $\rho$ when the size of the intervals
characterizing the uncertainties tends to zero, and the set of
single-point QIVPMs densely covers the entire real interval $[0,1]$.

The paper is organized as follows.  In Sec.\ \ref{sec:fuzzy} we begin
by introducing the foundations of quantum probability space and fuzzy
measurement.  We then move on to introduce the concept of quantum
interval valued probability in Sec.\ \ref{sec:Interval-Uncertainty}.
The QIVPM will provide the mathematical framework we need to quantify the
impact of finite precision measurements on quantum mechanics.  For
instance, in Sec.\ \ref{sec:Kochen-Specker}, we quantify the domain of
validity of a contextual measurement, thus addressing the conditions
under which the Kochen-Specker theorem applies.  This provides a way
not only of resolving the Meyer-Mermin debate, but also revealing a
precise transition to non-contextuality;  these are the results of our
Thm~\ref{cor:Kochen-Specker-IVPM}.  In Sec.\ \ref{sec:Gleason}, we
study Gleason's theorem in the context of imprecise measurements,
concluding that QIVPM provides a framework with quantitative bounds in
which Gleason's theorem, while formally invalid in a universe with
bounded resources, holds asymptotically in a mathematically precise
way.  Finally, we present our conclusions in Sec.~\ref{sec:Conclusion}.
